---
title: "Diffusion Modeling"
author: "William Hopper"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = 'svg')
library(rprojroot)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(optimx)
library(foreach)
library(rtdists)
library(dplyr)
root_dir <- rprojroot::is_rstudio_project$find_file()
```

```{css}
.main-container {
    max-width: 1200px;
}
img {
  padding-top: 10px;
}
```

```{r load_data}
load(file.path(root_dir, "data","SvDPRec.Rdata"))
test <- select(test, -list)
```

```{r cell_counts}
delayed_counts_by_sub_bias <- filter(test, !is.na(type)) %>%
  select(-starts_with("speeded")) %>%
  rename(correct = delayed_correct) %>%
  count(subject, strength, pOld, correct) %>%
  complete(subject, strength, pOld, correct,
           fill = list(n=0)) %>%
  unite(col = "resp_type", strength, correct) %>%
  spread(resp_type, n) %>%
  rename(FA = L_FALSE, CR = L_TRUE,
         M_S = S_FALSE, H_S = S_TRUE,
         M_W = W_FALSE, H_W = W_TRUE) %>%
  mutate(FA_N = FA + CR,
         S_N = M_S + H_S,
         W_N = M_W + H_W) %>%
  select(subject, pOld, FA, FA_N, H_S, S_N, H_W, W_N)

speeded_RT_choice_quantiles <- filter(test, !is.na(type)) %>%
  filter(speeded_RT > 200) %>%
  mutate(speeded_RT = speeded_RT/1000) %>%
  group_by(subject, strength, speeded_judgment) %>%
  summarise(quintiles = list(
              data.frame(quantile = c("10%", "30%", "50%", "70%", "90%", "100%"),
                         value = quantile(speeded_RT, c(.1, .3, .5, .7, .9, 1)),
                         bin_counts = floor(n() * c(.1, .2, .2, .2, .2, .1)),
                         stringsAsFactors = FALSE)
              )) %>%
  ungroup() %>%
  unnest()

speeded_acc <- filter(test, !is.na(type)) %>%
  group_by(subject, strength) %>%
  summarise(acc = mean(speeded_correct))

```

```{r SDT_model_functions}
SDT <- function(d, s, crit) {

  # if (is.unsorted(rev(crit))) {
  #   # warning("Criterion locations will be sorted in descending order")
  #   crit <- sort(crit, decreasing = TRUE)
  # }
  
  if (length(d) != 2L) {
    stop("Only 2 target distributions may be specified")
  }

  if (length(s) == 1L) {
    s <- c(s, s)
  }
  
  FAR <- pnorm(crit, lower.tail = FALSE)
  HR_W <- pnorm(crit, mean = d[1], sd = s[1], lower.tail = FALSE)
  HR_S <- pnorm(crit, mean = d[2], sd = s[2], lower.tail = FALSE)
  
  return(list("HitRate_Strong" = HR_S,
              "HitRate_Weak" = HR_W,
              "FalseAlarmRate" = FAR)
         )
}

SDT_LL <- function(H_S, H_W, FA, ...) {

  SDT_preds <- SDT(...)
  LL <- c(dbinom(H_S$H_S, size = H_S$S_N, prob = SDT_preds$HitRate_Strong, log=TRUE),
          dbinom(H_W$H_W, size = H_W$W_N, prob = SDT_preds$HitRate_Weak, log=TRUE),
          dbinom(FA$FA, size = FA$FA_N, prob = SDT_preds$FalseAlarmRate, log=TRUE)
          )
  return(-sum(LL))
}
```

```{r diffusion_functions, cache=TRUE}

SDT2diffusion <- function(parameters) {
  
    parameters['v.W'] <- parameters['v.L'] + parameters['v.W']*parameters['sv.L']
    parameters['v.S'] <- parameters['v.L'] + parameters['v.S']*parameters['sv.L']
    parameters['sv.T'] <- parameters['sv.T']*parameters['sv.L']
    return(parameters)
}
diffusion_LL <- function(parameters, data, fixed=NULL, scale=FALSE){

  parameters <- c(parameters, fixed)
  
  if (scale) {
    parameters <- SDT2diffusion(parameters)
  }
  
  log_likelihood <- 0

  all_strengths <- unique(data$strength) # Conditions to loop over
  boundaries <- 1:2 # Boundaries to loop over
  
  for (strength in all_strengths) {

    condition_index <- data$strength==strength
    bin_counts <- data$bin_counts[condition_index]
    p <- numeric(sum(condition_index))
    
    if (strength == "L") {
      sv = parameters["sv.L"]
    } else {
      sv = parameters["sv.T"]
    }

    insert_at = 0
    for (b in boundaries) {

      i <- condition_index & data$boundary == b
      cutpoints <- data$value[i]
      cutpoints[length(cutpoints)] <- Inf

      p_b <- pdiffusion(rt = cutpoints, response = b,
                        a = parameters["a"],
                        z = parameters["a"] * parameters["z"],
                        t0 = parameters["t0"],
                        st0 = parameters["st0"],
                        v = parameters[paste0("v.", strength)],
                        sv = sv)

      p_b <- c(p_b[1], diff(p_b))
      less <- p_b <= 0
      if (any(less)) {
        p_b[less] <- 0.0001
        p_b <- p_b/sum(p_b)
      }

      p[(insert_at+1):(insert_at+length(p_b))] <- p_b
      insert_at <- insert_at + length(p_b)
    }

    log_likelihood <- log_likelihood + dmultinom(x = bin_counts,
                                                 size = sum(bin_counts),
                                                 prob = p,
                                                 log = TRUE)
    if (is.infinite(log_likelihood)) {
      return(1e9)
    }
  }

  return(-log_likelihood)

}
```

```{r start_cluster}
cl <- parallel::makeCluster(parallel::detectCores(),
                            outfile = "",
                            methods = FALSE)

doParallel::registerDoParallel(cl)
```

## Diffusion Model Parameters
- a = boundary separation
- z = starting point (parameterized here as a proportion of boundary separation)
- t0 = lower bound of non-decision RT component
- st0 = range of uniform distribution of non-decision times. NDT distribution is U(t0, t0+st0)
- v.L = drift rate for lures (fixed at 1)
- v.W = drift rate for weak targets
- v.S = drift rate for strong targets
- sv.L = inter-trial drift variance for lures
- cv.T = inter-trial drift variance for targets (both weak and strong)

## UVSD Model Paramters
- d.W = Mean of "weak" target distribution
- d.S = Mean of "strong" target distribution
- s = Varaibility of both target distributions 
- crit1 = Threshold for conservative bias condition
- crit2 = Threshold for neutral bias condition
- crit3 = Threshold for weak bias condition

```{r fit, cache=TRUE}
SDT_diffusion_LL_sum <- function(theta, ...) {
  
  trailing <- list(...)
  
  
  
  SDT_nLL <- SDT_LL(d = theta[c("d.W","d.S")],
                    s = theta["s"],
                    crit = theta[c("crit1", "crit2", "crit3")],
                    H_S = trailing$H_S,
                    H_W = trailing$H_W,
                    FA = trailing$FA
                    )

  diff_nLL <- diffusion_LL(theta[c("a", "z", "t0", "st0", "v.L", "sv.L")],
                           data = trailing$data,
                           fixed = c("v.W" = unname(theta["d.W"]),
                                     "v.S" = unname(theta["d.S"]),
                                     "sv.T" = unname(theta["s"])),
                           scale = TRUE
                           )

  negLL <- SDT_nLL + diff_nLL

  if (is.infinite(negLL) || is.nan(negLL)) {
    negLL <- .Machine$double.xmax
  }
  negLL
}

subjectwise_datalist <- Map(
  function(x,y) { list("speeded" = x,"biased" = y)},
  split(speeded_RT_choice_quantiles, speeded_RT_choice_quantiles$subject),
  split(delayed_counts_by_sub_bias, delayed_counts_by_sub_bias$subject)
  )

fits <- foreach(sub = subjectwise_datalist,
                .packages = c("rtdists","optimx"))  %dopar% {

  # Make incorrects the lower boundary responses (1) and 
  # corrects the upper boundary responses(2).
  # This will be useful inside the objective function
  sub$speeded$boundary <- as.numeric(
    factor(sub$speeded$speeded_judgment,
           levels = c('n','o'),
           labels = c("lower",'upper')
           )
    )
  
  theta <- c("d.W" = 1.5, "d.S" = 2.5, "s" = 1.2,
             crit = c(1.25, 1, .75),
             "a" = 1, "z" = .5, "t0" = .1, "st0" = .1,
             "v.L" = 0, "sv.L" = .5
             )
  fit <- optimx(par = theta,
                fn = SDT_diffusion_LL_sum,
                method = "nlminb",
                itnmax = 3000,
                lower = c("d.W" = 0, "d.S" = 0, "s" = 0.1,
                        crit = c(-3, -3, -3),
                        "a" = .01, "z" = 0.5, "t0" = .01, "st0" = .01,
                        "v.L" = -5, "sv.L" = .1
                        ),
                upper = c("d.W" = 5, "d.S" = 5, "s" = 2,
                        crit = c(3, 3, 3),
                        "a" = 3, "z" = .95, "t0" = .5, "st0" = .5,
                        "v.L" = 5, "sv.L" = 4
                        ),
                control = list(kkt=FALSE),
                data = sub$speeded, # passed to diffusion_LL
                H_S = sub$biased[c("H_S","S_N")], # passed to SDT_LL
                H_W = sub$biased[c("H_W", "W_N")],# passed to SDT_LL
                FA = sub$biased[c("FA", "FA_N")] # passed to SDT_LL
              )

  fit$subject <- sub$speeded$subject[1]
  fit[c("v.W","v.S","sv.T")] <- fit[c("d.W","d.S","s")]  # This looks stupid but is useful
  scaled <- SDT2diffusion(unlist(fit[c('v.L','sv.L',"v.W","v.S","sv.T")]))
  fit[c("v.W","v.S","sv.T")] <- as.list(scaled[c("v.W","v.S","sv.T")])
  fit
}

fits <- bind_rows(fits)
```

```{r stop_cluster}
parallel::stopCluster(cl)
```

```{r param_processing}
SDT_params <- select(fits, subject, d.W:crit3)
diffusion_params <- select(fits, subject, v.L, sv.L,v.W, v.S, sv.T, a:st0)
GOF <- select(fits, subject, value:convcode)
```

```{r SDT_preds, fig.height=12, fig.width=9}
FAR_points <- seq(0, 1, .01)
SDT_preds <- rowwise(SDT_params) %>%
  mutate(SDT = list(
    data.frame(FAR = rep(FAR_points, 2),
               strength = rep(c("W", "S"), each = length(FAR_points)),
               HR = pnorm(c(qnorm(FAR_points)/s + d.W,
                            qnorm(FAR_points)/s + d.S
                            )
                          ),
               type = "pred")
    )) %>%
  select(subject, SDT) %>%
  unnest()

ROC_by_sub <- mutate(delayed_counts_by_sub_bias,
         FAR = FA / FA_N,
         HR_W = H_W / W_N,
         HR_S = H_S / S_N
         ) %>%
  select(subject, pOld, FAR, HR_W, HR_S) %>%
  gather(key = "strength", value = "HR", HR_W, HR_S) %>%
  arrange(subject, desc(strength)) %>%
  mutate(strength = sub("HR_", "", strength, fixed=TRUE))

ggplot(SDT_preds,
       aes(x=FAR, y=HR, color=strength)) +
  geom_line(size=.8) +
  geom_point(aes(shape = factor(pOld)),
             data=ROC_by_sub) +
  facet_wrap(~ subject,
             labeller = labeller(subject = label_both,
                                 .multi_line = FALSE),
             ncol=4) +
  scale_x_continuous("False Alarm Rate", limits = c(0,1),
                     labels = c(0,0.25, 0.5, 0.75, 1)) +
  scale_y_continuous("Hit Rate", limits = c(0,1)) +
  scale_color_manual("Strength", values = c("#2470b7", "#b70e1c")) +
  scale_shape_discrete("pOld") +
  guides(colour = guide_legend(override.aes = list(size = 4)),
         shape = guide_legend(override.aes = list(size = 4))) +
  coord_fixed() +
  ggtitle("Predicted ROC for delayed judgments") +
  theme_bw(base_size = 13) +
  theme(legend.position = c(.75, .1),
        legend.box = "horizontal")
```

