---
title: "Diffusion Modeling"
author: "William Hopper"
date: "`r format(Sys.Date(), '%b %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = 'svg')
library(rprojroot)
library(tidyr)
library(ggplot2)
library(kableExtra)
library(optimx)
library(foreach)
library(rtdists)
library(dplyr)
root_dir <- rprojroot::is_rstudio_project$find_file()
```

```{css}
.main-container {
    max-width: 1200px;
}
img {
  padding-top: 10px;
}
```


```{r load_data}
load(file.path(root_dir, "data","SvDPRec.Rdata"))
test <- select(test, -list)
```


```{r SDT_model_functions, cache=TRUE, dependson=c("cell_counts")}
SDT <- function(d, s, crit) {
  
  if (length(d) != 2L) {
    stop("Only 2 target distributions may be specified")
  }

  if (length(s) == 1L) {
    s <- c(s, s)
  }
  
  FAR <- pnorm(crit, lower.tail = FALSE)
  HR_W <- pnorm(crit, mean = d[1], sd = s[1], lower.tail = FALSE)
  HR_S <- pnorm(crit, mean = d[2], sd = s[2], lower.tail = FALSE)
  
  return(list("HitRate_Strong" = HR_S,
              "HitRate_Weak" = HR_W,
              "FalseAlarmRate" = FAR)
         )
}

SDT_LL <- function(H_S, H_W, FA, ...) {

  SDT_preds <- SDT(...)
  LL <- c(dbinom(H_S$H_S, size = H_S$S_N, prob = SDT_preds$HitRate_Strong, log=TRUE),
          dbinom(H_W$H_W, size = H_W$W_N, prob = SDT_preds$HitRate_Weak, log=TRUE),
          dbinom(FA$FA, size = FA$FA_N, prob = SDT_preds$FalseAlarmRate, log=TRUE)
          )
  return(-sum(LL))
}
```


```{r diffusion_functions, cache=TRUE, dependson=c("cell_counts")}

SDT2diffusion <- function(parameters) {
  
    parameters['v.W'] <- parameters['v.L'] + parameters['d.W']*parameters['sv.L']
    parameters['v.S'] <- parameters['v.L'] + parameters['d.S']*parameters['sv.L']
    parameters['sv.T'] <- parameters['s']*parameters['sv.L']
    return(parameters)
}

diffusion_LL <- function(parameters, data, fixed=NULL){

  parameters <- c(parameters, fixed)
  
  log_likelihood <- 0

  all_strengths <- unique(data$strength) # Conditions to loop over
  boundaries <- 1:2 # Boundaries to loop over
  
  for (strength in all_strengths) {

    condition_index <- data$strength==strength
    bin_counts <- data$bin_counts[condition_index]
    p <- numeric(sum(condition_index))
    
    if (strength == "L") {
      sv = parameters["sv.L"]
    } else {
      sv = parameters["sv.T"]
    }

    insert_at = 0
    for (b in boundaries) {

      i <- condition_index & data$boundary == b
      cutpoints <- data$value[i]
      cutpoints[length(cutpoints)] <- Inf

      p_b <- pdiffusion(rt = cutpoints, response = b,
                        a = parameters["a"],
                        z = parameters["a"] * parameters["z"],
                        t0 = parameters["t0"],
                        st0 = parameters["st0"],
                        v = parameters[paste0("v.", strength)],
                        sv = sv)

      p_b <- c(p_b[1], diff(p_b))
      less <- p_b <= 0
      if (any(less)) {
        p_b[less] <- 0.0001
        p_b <- p_b/sum(p_b)
      }

      p[(insert_at+1):(insert_at+length(p_b))] <- p_b
      insert_at <- insert_at + length(p_b)
    }

    log_likelihood <- log_likelihood + dmultinom(x = bin_counts,
                                                 size = sum(bin_counts),
                                                 prob = p,
                                                 log = TRUE)
    if (is.infinite(log_likelihood)) {
      return(1e9)
    }
  }

  return(-log_likelihood)

}
```


# By-Subject Modeling

```{r cell_counts, cache=TRUE}
delayed_counts_by_sub_bias <- filter(test, !is.na(type)) %>%
  select(-starts_with("speeded")) %>%
  rename(correct = delayed_correct) %>%
  count(subject, strength, pOld, correct) %>%
  complete(subject, strength, pOld, correct,
           fill = list(n=0)) %>%
  unite(col = "resp_type", strength, correct) %>%
  spread(resp_type, n) %>%
  rename(FA = L_FALSE, CR = L_TRUE,
         M_S = S_FALSE, H_S = S_TRUE,
         M_W = W_FALSE, H_W = W_TRUE) %>%
  mutate(FA_N = FA + CR,
         S_N = M_S + H_S,
         W_N = M_W + H_W) %>%
  select(subject, pOld, FA, FA_N, H_S, S_N, H_W, W_N)

speeded_RT_choice_quantiles <- filter(test, !is.na(type)) %>%
  filter(speeded_RT > 200) %>%
  mutate(speeded_RT = speeded_RT/1000) %>%
  group_by(subject, strength, speeded_judgment) %>%
  summarise(quintiles = list(
              data.frame(quantile = c("10%", "30%", "50%", "70%", "90%", "100%"),
                         value = quantile(speeded_RT, c(.1, .3, .5, .7, .9, 1)),
                         bin_counts = floor(n() * c(.1, .2, .2, .2, .2, .1)),
                         stringsAsFactors = FALSE)
              )) %>%
  ungroup() %>%
  unnest()

speeded_acc <- filter(test, !is.na(type)) %>%
  group_by(subject, strength) %>%
  summarise(acc = mean(speeded_correct))

```


### Diffusion Model Parameters
- `a` = boundary separation
- `z` = starting point (parameterized here as a proportion of boundary separation)
- `t0` = lower bound of non-decision RT component
- `st0` = range of uniform distribution of non-decision times. NDT distribution is U(t0, t0+st0)
- `v.L` = drift rate for lures
- `v.W` = drift rate for weak targets
- `v.S` = drift rate for strong targets
- `sv.L` = inter-trial drift variance for lures
- `cv.T` = inter-trial drift variance for targets (both weak and strong)

## UVSD Model Paramters
- `d.W` = Mean of "weak" target distribution
- `d.S` = Mean of "strong" target distribution
- `s` = Varaibility of both target distributions 
- `crit1` = Threshold for conservative bias condition
- `crit2` = Threshold for neutral bias condition
- `crit3` = Threshold for weak bias condition


```{r starting_theta}
starting_theta <- tribble(
  ~subject,    ~a,    ~z,  ~t0,    ~st0,   ~v.L, ~sv.L, ~d.W,  ~d.S,   ~s, ~crit1, ~crit2, ~crit3,
  "101",    0.989, 0.612, 0.342, 0.1848, -1.576, 1.167, 1.259, 2.630, 1.361, 1.273, 0.783, 0.293,
  "102",    1.165, 0.621, 0.367, 0.2444, -0.414, 0.735, 0.699, 1.266, 1.036, 0.649, 0.189,-0.070,
  "103",    0.846, 0.620, 0.399, 0.0365, -1.523, 0.406, 1.641, 2.528, 1.220, 1.200, 0.646, 0.004,
  "104",    2.214, 0.532, 0.420, 0.0500, -2.595, 2.212, 1.826, 2.755, 1.172, 1.681, 1.123, 0.754,
  "105",    0.915, 0.702, 0.244, 0.2873,  0.527, 1.594, 0.610, 1.531, 0.967, 1.274,-0.403,-1.217,
  "106",    1.258, 0.560, 0.299, 0.2534, -1.367, 0.774, 1.489, 3.309, 1.509, 1.237, 0.825, 0.488,
  "107",    1.197, 0.562, 0.439, 0.2175, -1.861, 1.112, 1.396, 2.287, 1.167, 1.527, 0.800, 0.346,
  "108",    0.939, 0.514, 0.392, 0.0974, -2.428, 0.309, 2.465, 3.950, 0.941, 1.894, 1.604, 1.203,
  "109",    1.243, 0.656, 0.432, 0.0372, -2.646, 0.468, 2.547, 3.950, 1.214, 1.642, 1.541, 1.237,
  "110",    0.840, 0.627, 0.335, 0.2182, -1.584, 0.237, 1.165, 2.236, 0.870, 1.407, 0.988, 0.842,
  "111",    0.947, 0.686, 0.432, 0.0944, -2.354, 0.214, 1.686, 3.552, 1.757, 1.406, 0.956, 0.209,
  "112",    1.215, 0.630, 0.436, 0.1639, -1.222, 0.791, 0.987, 1.627, 0.586, 0.725, 0.674, 0.597,
  "113",    1.259, 0.516, 0.483, 0.1166, -2.883, 1.006, 2.318, 3.950, 0.998, 1.994, 1.482, 1.436,
  "114",    1.017, 0.615, 0.293, 0.2657, -1.082, 0.214, 0.995, 1.585, 0.956, 1.279, 0.720, 0.050,
  "101",    0.998, 0.617, 0.342, 0.1853, -1.631, 1.369, 1.406, 3.020, 1.647, 1.291, 0.798, 0.259,
  "102",    1.250, 0.607, 0.361, 0.2488, -0.284, 1.481, 0.650, 1.140, 0.849, 0.638, 0.217, 0.004,
  "103",    0.770, 0.672, 0.373, 0.1635, -2.267, 1.690, 1.858, 3.000, 1.422, 1.232, 0.669,-0.003,
  "104",    1.566, 0.517, 0.412, 0.013,  -0.885, 0.839, 1.579, 2.610, 1.110, 1.414, 0.994, 0.667,
  "105",    0.897, 0.705, 0.245, 0.2900,  0.333, 0.943, 0.571, 1.520, 0.922, 1.265,-0.397,-1.186,
  "106",    1.304, 0.563, 0.298, 0.2661, -1.465, 1.153, 1.725, 3.390, 1.540, 1.261, 0.870, 0.555,
  "107",    1.227, 0.573, 0.438, 0.2248, -2.074, 1.531, 1.644, 2.840, 1.562, 1.614, 0.794, 0.313,
  "108",    0.846, 0.500, 0.414, 0.0198, -0.268, 0.805, 1.644, 2.660, 0.995, 1.307, 1.008, 0.707,
  "109",    1.033, 0.555, 0.405, 0.0911, -0.303, 0.916, 1.634, 2.660, 1.047, 1.271, 1.004, 0.733,
  "110",    0.864, 0.638, 0.335, 0.2203, -1.703, 1.103, 1.250, 2.630, 1.136, 1.454, 1.000, 0.828,
  "111",    1.002, 0.711, 0.432, 0.1018, -2.816, 1.594, 1.884, 3.980, 1.900, 1.425, 0.975, 0.205,
  "112",    1.279, 0.634, 0.434, 0.1725, -1.322, 1.388, 1.435, 3.050, 1.305, 0.754, 0.702, 0.611,
  "113",    1.206, 0.500, 0.450, 0.0396, -1.194, 1.124, 1.626, 3.000, 0.839, 1.516, 1.104, 0.901,
  "114",    1.081, 0.646, 0.294, 0.2683, -1.460, 1.600, 1.083, 2.100, 1.452, 1.351, 0.721,-0.119
)
```


```{r fit, cache=TRUE, dependson=c("diffusion_functions", "SDT_model_functions")}
cl <- parallel::makeCluster(parallel::detectCores(),
                            outfile = "",
                            methods = FALSE)

doParallel::registerDoParallel(cl)

SDT_diffusion_LL_sum <- function(theta, data, theta_names, scale = TRUE) {

  names(theta) <- theta_names
  SDT_nLL <- SDT_LL(d = theta[c("d.W","d.S")],
                    s = theta["s"],
                    crit = theta[c("crit1", "crit2", "crit3")],
                    H_S = data$biased[c("H_S","S_N")],
                    H_W = data$biased[c("H_W", "W_N")],
                    FA = data$biased[c("FA", "FA_N")]
                    )

  if (scale) {
    theta <- SDT2diffusion(theta)
  }
  
  diff_nLL <- diffusion_LL(theta[c("a", "z", "t0", "st0", "v.L", "sv.L", "v.W", "v.S", "sv.T")],
                           data = data$speeded
                           )

  negLL <- SDT_nLL + diff_nLL

  if (is.infinite(negLL) || is.nan(negLL)) {
    negLL <- .Machine$double.xmax
  }
  negLL
}

subjectwise_datalist <- Map(
  function(x,y) { list("speeded" = x, "biased" = y)},
  split(speeded_RT_choice_quantiles, speeded_RT_choice_quantiles$subject),
  split(delayed_counts_by_sub_bias, delayed_counts_by_sub_bias$subject)
  )

fits <- foreach(sub = subjectwise_datalist,
                .packages = c("rtdists","optimx", "tibble"))  %dopar% {

  # Make incorrects the lower boundary responses (1) and 
  # corrects the upper boundary responses(2).
  # This will be useful inside the objective function
  sub$speeded$boundary <- as.numeric(
    factor(sub$speeded$speeded_judgment,
           levels = c('n','o'),
           labels = c("lower",'upper')
           )
    )
  
  x <- starting_theta[starting_theta$subject == sub$speeded$subject[1], -1]
  theta <- list(unlist(x[1,]), unlist(x[2,]),
                c("a" = 1, "z" = .5, "t0" = .1, "st0" = .1, "v.L" = 0, "sv.L" = .5,
                  "d.W" = 1.5, "d.S" = 2.5, "s" = 1.2, crit = c(1.25, 1, .75))
                )
  sub_fits <- list()
  
  for (i in 1:length(theta)) {
    fit <- optimx(par = theta[[i]],
                  fn = SDT_diffusion_LL_sum,
                  method = c("nmkb", "nlminb"),
                  itnmax = 3000,
                  lower = c("a" = .01, "z" = 0.05, "t0" = .01, "st0" = .01,
                            "v.L" = -5, "sv.L" = .1,
                            "d.W" = 0, "d.S" = 0, "s" = 0.1,
                            crit = c(-3, -3, -3)
                            ),
                  upper = c("a" = 3, "z" = .95, "t0" = .5, "st0" = .5,
                            "v.L" = 5, "sv.L" = 4,
                            "d.W" = 4, "d.S" = 4, "s" = 2,
                            crit = c(3, 3, 3)
                            ),
                  control = list(kkt=FALSE, follow.on=FALSE),
                  data = sub,
                  theta_names = names(theta[[i]])
                  )
      fit$subject <- sub$speeded$subject[1]
      fit <- tibble::rownames_to_column(fit, "method")
      for (j in nrow(fit)) {
        scaled <- SDT2diffusion(parameters = unlist(fit[j, c("d.W", "d.S", "s", "sv.L", "v.L")]))
        fit[c("v.W","v.S","sv.T")] <- as.list(scaled[c("v.W","v.S","sv.T")])
      }
      sub_fits[[i]] <- fit
  }

  sub_fits

}

fits <- lapply(fits, bind_rows) %>%
  bind_rows() %>%
  group_by(subject) %>%
  filter(value == min(value)) %>%
  ungroup()

parallel::stopCluster(cl)
```

```{r param_processing, cache=TRUE, dependson="fit"}
GOF <- select(fits, subject, value:convcode) %>%
  rename(`-LL` = value) %>%
  mutate(deviance = 2*`-LL`)

SDT_params <- select(fits, subject, d.W:crit3)

diffusion_params <- select(fits, subject, v.L, sv.L,v.W, v.S, sv.T, a:st0) %>%
  mutate(sv.W = sv.T) %>%
  rename(sv.S = sv.T) %>%
  gather(key='param', value='value', v.L, v.W, v.S, sv.L, sv.W, sv.S) %>%
  separate(param, into=c("param", "strength")) %>%
  spread(param, value) %>%
  select(subject, strength, a:st0, v, sv)

```

```{r SDT_sub_preds, fig.height=12, fig.width=9, cache=TRUE, dependson="fit"}
FAR_points <- seq(0, 1, .01)
SDT_preds <- rowwise(SDT_params) %>%
  mutate(SDT = list(
    data.frame(FAR = rep(FAR_points, 2),
               strength = rep(c("W", "S"), each = length(FAR_points)),
               HR = pnorm(c((qnorm(FAR_points) + d.W)/s,
                            (qnorm(FAR_points) + d.S)/s
                            )
                          ),
               type = "pred")
    )) %>%
  select(subject, SDT) %>%
  unnest()


ROC_by_sub <- mutate(delayed_counts_by_sub_bias,
         FAR = FA / FA_N,
         HR_W = H_W / W_N,
         HR_S = H_S / S_N
         ) %>%
  select(subject, pOld, FAR, HR_W, HR_S) %>%
  gather(key = "strength", value = "HR", HR_W, HR_S) %>%
  arrange(subject, desc(strength)) %>%
  mutate(strength = sub("HR_", "", strength, fixed=TRUE))

ggplot(SDT_preds,
       aes(x=FAR, y=HR, color=strength)) +
  geom_line(size=.8) +
  geom_point(aes(shape = factor(pOld)),
             data=ROC_by_sub) +
  facet_wrap(~ subject,
             labeller = labeller(subject = label_both,
                                 .multi_line = FALSE),
             ncol=4) +
  scale_x_continuous("False Alarm Rate", limits = c(0,1),
                     labels = c(0,0.25, 0.5, 0.75, 1)) +
  scale_y_continuous("Hit Rate", limits = c(0,1)) +
  scale_color_manual("Strength", values = c("#2470b7", "#b70e1c")) +
  scale_shape_discrete("pOld") +
  guides(colour = guide_legend(override.aes = list(size = 4)),
         shape = guide_legend(override.aes = list(size = 4))) +
  coord_fixed() +
  ggtitle("Predicted ROC for delayed judgments") +
  theme_bw(base_size = 13) +
  theme(legend.position = c(.75, .1),
        legend.box = "horizontal")
```

```{r SDT_avg_preds, fig.height=8, fig.width=8, cache=TRUE, dependson="fit"}
SDT_preds_avg <-
  summarise(.data = data.frame(t(colMeans(SDT_params[,2:7]))),
            SDT = list(data.frame(
                FAR = rep(FAR_points, 2),
                HR = pnorm(c((qnorm(FAR_points) + d.W)/s,
                             (qnorm(FAR_points) + d.S)/s
                           )),
                strength = factor(rep(c("W", "S"), each = length(FAR_points)),
                                levels=c("W","S")
                                )
                ))
            )%>%
  unnest()

avg_obs_ROC_data <- group_by(ROC_by_sub, pOld, strength) %>%
  summarise_at(.vars = c("FAR", "HR"), .funs = "mean") %>%
  ungroup()

ggplot(SDT_preds_avg,
       aes(x=FAR, y=HR, linetype=strength)) +
  geom_line(size=1.75) +
  geom_point(aes(color=factor(pOld)),
             data = avg_obs_ROC_data,
             size=4) +
  scale_x_continuous("False Alarm Rate", limits = c(0,1),
                     labels = c(0,0.25, 0.5, 0.75, 1)) +
  scale_y_continuous("Hit Rate", limits = c(0,1)) +
  scale_linetype_discrete("Item",
                          labels=c("W" = "'Weak' Target",
                                   "S" = "'Strong' Target")) +
  scale_color_manual("p(Old)", values = c("#d82d2d", "#d8d834", "#17b50c")) +
  guides(colour = guide_legend(override.aes = list(size = 4)),
         shape = guide_legend(override.aes = list(size = 4))) +
  coord_fixed() +
  ggtitle("UVSD Model ROC (Average)") +
  theme_bw(base_size = 16) +
  theme(plot.title = element_text(hjust=.5),
        strip.background = element_blank(),
        legend.key.width = unit(10,"mm")
        )
```

```{r diffusion_predictions, cache=TRUE, dependson="fit"}
diffusion_acc_preds <- rowwise(diffusion_params) %>%
  mutate(p_resp_old = pdiffusion(Inf, response = 2, a=a, v=v, t0=t0, z=z*a, sv=sv, st0=st0),
         acc = replace(p_resp_old, strength == "L", 1 - p_resp_old[strength == "L"])
         ) %>%
  select(-p_resp_old) %>%
  ungroup()

accuracy <- bind_rows(predicted = select(mutate(diffusion_acc_preds, model="diffusion"),
                                         subject, model, strength, acc),
                      observed = mutate(speeded_acc, model = 'observed')
                      )

diffusion_RT_quantiles <- rowwise(diffusion_params) %>%
  mutate(RT = list(as.data.frame(matrix(
                qdiffusion(p = rep(c(.1, .3, .5, .7, .9), 2),
                           response = rep(1:2, each = 5),
                           a = a, z = z*a, t0 = t0, st0 = st0,
                           v = v, sv = sv,
                           scale_p = TRUE),
                nrow=2, ncol=5,
                byrow=TRUE,
                dimnames = list(c("Incorrect","Correct"),
                                quantile = c("10%","30%", "50%", "70%", "90%"))
                )))
         ) %>%
  unnest() %>% 
  mutate(speeded_judgment = rep(c('n', 'o'), nrow(.)/2)) %>%
  select(subject, strength, speeded_judgment, `10%`, `30%`, `50%`, `70%`, `90%`) %>%
  gather('quantile', 'value', `10%`, `30%`, `50%`, `70%`, `90%`)

RT_quantiles <- bind_rows(predicted = mutate(diffusion_RT_quantiles, model="diffusion"),
                          observed = filter(speeded_RT_choice_quantiles, quantile != "100%") %>%
                                     select(-bin_counts) %>%
                                     mutate(model = 'observed')
                          ) %>%
  mutate(speeded_correct = ifelse((strength == "L" & speeded_judgment == "n") |
                                    strength %in% c("S", "W") & speeded_judgment == "o",
                                  TRUE,
                                  FALSE)
         )

conditional_quantiles <- left_join(RT_quantiles,
                                   mutate(accuracy, inc = 1-acc) %>%
                                     gather("acc", "p", acc, inc) %>%
                                     mutate(speeded_correct = ifelse(acc=="acc",TRUE, FALSE)) %>%
                                     select(subject, model, strength, speeded_correct, p),
                                   by=c("subject", "model", "strength", "speeded_correct")) %>%
  select(-speeded_judgment) %>%
  rename(RT = value) %>%
  mutate(strength = factor(strength, levels=c("L","W","S"),
                           labels = c("Lure", "'Weak' Target", "'Strong' Target")),
         quantile = as.numeric(sub("%", "", quantile, fixed = TRUE))/100)

sub_conditional_quantiles <-  mutate(conditional_quantiles,
                                     quantile = quantile * p) %>%
  select(-p)

avg_conditional_quantiles <- group_by(conditional_quantiles,
                                      model, strength, quantile, speeded_correct) %>%
  summarise_at(c("RT","p"), mean) %>%
  ungroup() %>%
  mutate(quantile = quantile * p) %>%
  select(-p)
```

```{r QP_plots, fig.width=10, fig.height=7}
ggplot(avg_conditional_quantiles,
       aes(x=RT, y=quantile, color=speeded_correct, linetype=model, shape=model)) +
  geom_point(size=2) +
  geom_line(size=1) +
  facet_grid(~strength) +
  scale_color_manual("Accuracy",
                     values = RColorBrewer::brewer.pal(3, 'Set1')[1:2],
                     breaks=c("FALSE","TRUE"),
                     labels=c("Incorrect", "Correct")) +
  scale_linetype_discrete(NULL, labels = c("diffusion" = "Diffusion", "observed" = "Observed")) +
  ggtitle("UVSD + Diffusion Model",
          subtitle="Conditional Quantiles (Average)") +
  scale_x_continuous("Response Time (s.)") +
  guides(shape = FALSE) +
  theme_bw(base_size=16) +
  theme(legend.key.size = unit(15,"mm"),
        plot.title = element_text(hjust=.5),
        plot.subtitle = element_text(hjust=.5),
        strip.background = element_blank())
```


```{r g2, cache=TRUE, dependson="fit"}
speeded_RT_deviance <- speeded_RT_choice_quantiles %>%
  group_by(subject, strength) %>%
  mutate(observed_p = bin_counts/sum(bin_counts)) %>%
  summarise(LL = dmultinom(bin_counts, sum(bin_counts), observed_p, log=TRUE)) %>%
  group_by(subject) %>%
  summarise(null_deviance = -2*sum(LL))

delayed_acc_deviance <- mutate(delayed_counts_by_sub_bias,
         FA_LL = dbinom(FA, FA_N, FA/FA_N, log=TRUE),
         HW_LL = dbinom(H_W, W_N, H_W/W_N, log=TRUE),
         HS_LL = dbinom(H_S, S_N, H_S/S_N, log=TRUE),
         LL = FA_LL + HW_LL + HS_LL) %>%
  select(subject, pOld, LL) %>%
  group_by(subject) %>%
  summarise(null_deviance = -2*sum(LL))

null_deviance <- left_join(speeded_RT_deviance, delayed_acc_deviance,
                           by="subject",suffix=c(".RT",".acc")) %>%
  transmute(subject,
            null_deviance = null_deviance.RT + null_deviance.acc)

data_df <- 9 + 33 # 9 Hit/FA rates, 33 free RT bins
model_df <- 12 # 6 SDT  params, 9 diffusion params - 3 shared (the target drift and drift variability params)

g2 <- left_join(select(GOF, subject, deviance),
                null_deviance,
                by = "subject") %>% 
  mutate(G2 = deviance - null_deviance,
         p = pchisq(G2, df = data_df - model_df, lower.tail = FALSE)
         )

GOF <- left_join(select(GOF, -deviance),
                 select(g2, subject, G2, p),
                 by="subject")
```

```{r plots, results='asis', fig.height=6, fig.width=9, cache=TRUE, dependson="fit"}
for (s in unique(fits$subject)) {
  
  cat(sprintf("<h3>Subject %s</h3>", s))
  
  subject_params <- filter(diffusion_params, subject == s) %>%
    mutate(strength = factor(strength, levels = c("L", "W", "S"))) %>%
    arrange(strength)
  subject_GOF <- filter(GOF, subject == s)
  
  kable(subject_params, digits = 3) %>%
    kableExtra::kable_styling(full_width = FALSE, position = "float_left") %>%
    column_spec(1:2, bold=TRUE) %>%
    collapse_rows(c(1,3:6, 8)) %>%
    print()
  
  kable(subject_GOF, digits = 3) %>%
    kableExtra::kable_styling(full_width = FALSE, position = "float_right") %>%
    print()
  
  acc_plot <-
    ggplot(filter(accuracy, subject == s),
          aes(x=strength, y=acc, shape=model, color=strength)) +
    geom_point(size=2,stroke=2, position = position_dodge(.25)) +
    scale_shape_manual("",
                       values = c(16,4),
                       limits = c("observed","diffusion"),
                       labels = c("Observed", "Diffusion")) +
    scale_color_discrete(labels = c("L" = "Lure",
                                    "W" = "Weak\nTarget",
                                    "S" = "Strong\nTarget"),
                         breaks = c("L","W","S")) +
    scale_y_continuous("Percent Correct", limits = c(0, 1)) +
    scale_x_discrete(labels = c("L" = "Lure",
                                "W" = "Weak\nTarget",
                                "S" = "Strong\nTarget"),
                     limits = c("L","W","S"),
                     expand = c(0, .2)) +
    theme(legend.key.height = unit(9, 'mm'))
  
  quantile_plot <- 
    ggplot(filter(RT_quantiles, subject == s),
           aes(x=quantile, y=value, shape=model, color=strength)) +
    geom_point(size=2) +
    geom_line(aes(group=model)) +
    facet_grid(speeded_correct ~ strength,
               labeller = labeller(speeded_correct = as_labeller(
                                    c("TRUE" = "Correct",
                                      "FALSE" = "Error")
                                    ),
                                    strength = as_labeller(
                                      c("L" = "Lure",
                                        "S" = "Strong Target",
                                        "W" = "Weak Target")
                                    )
                                   )
               ) +
    scale_shape_manual("",
                       values = c(16,4),
                       limits = c("observed", "diffusion"),
                       labels = c("Observed", "Diffusion")) +
    scale_y_continuous("RT Quantile Value (s.)") +
    guides(color=FALSE, shape=FALSE)
  
  gridExtra::grid.arrange(acc_plot, quantile_plot, nrow=1,
                          widths=c(.35, .65))
}
```

# Aggregated Modeling

```{r pooled_data}
delayed_counts_by_bias <-
  select(.data = delayed_counts_by_sub_bias, -subject) %>%
  group_by(pOld) %>%
  summarise_all(.funs = "sum")

obs_ROC_data_pooled <-
  mutate(delayed_counts_by_bias,
         FAR = FA / L_N,
         HR_W = H_W / W_N,
         HR_S = H_S / S_N,
         pOld = sprintf("%2.0f%%", pOld * 100)) %>%
  select(pOld, FAR, HR_W, HR_S) %>%
  gather(key="strength", value="HR", HR_W, HR_S) %>%
  mutate(strength = sub("HR_", "", strength)) %>%
  select(pOld, strength, FAR, HR)

pooled_RT_choice_quantiles <-
  group_by(.data = speeded_RT_choice_quantiles, strength, speeded_judgment, quantile) %>%
  summarise(value = mean(value),
            bin_counts = sum(bin_counts)
            ) %>%
  ungroup() %>%
  arrange(strength, speeded_judgment, value) %>%
  group_by(strength, speeded_judgment) %>%
  mutate(bin_counts = floor(bin_counts * c(.1, .2, .2, .2, .2, .1))) %>%
  ungroup()

pooled_acc <- filter(test, !is.na(type), speeded_RT > 200) %>%
  group_by(strength) %>%
  summarise(acc = mean(speeded_correct))
```


```{r fit_pooled}
theta <- list(unlist(x[1,]), unlist(x[2,]),
              c("a" = 1, "z" = .5, "t0" = .1, "st0" = .1, "v.L" = 0, "sv.L" = .5,
                "d.W" = 1.5, "d.S" = 2.5, "s" = 1.2, crit = c(1.25, 1, .75))
              )

fits <- list()
  
  for (i in 1:length(theta)) {
    fit <- optimx(par = theta[[i]],
                  fn = SDT_diffusion_LL_sum,
                  method = c("nmkb", "nlminb"),
                  itnmax = 3000,
                  lower = c("a" = .01, "z" = 0.05, "t0" = .01, "st0" = .01,
                            "v.L" = -5, "sv.L" = .1,
                            "d.W" = 0, "d.S" = 0, "s" = 0.1,
                            crit = c(-3, -3, -3)
                            ),
                  upper = c("a" = 3, "z" = .95, "t0" = .5, "st0" = .5,
                            "v.L" = 5, "sv.L" = 4,
                            "d.W" = 4, "d.S" = 4, "s" = 2,
                            crit = c(3, 3, 3)
                            ),
                  control = list(kkt=FALSE, follow.on=FALSE),
                  data = sub,
                  theta_names = names(theta[[i]])
                  )
      fit$subject <- sub$speeded$subject[1]
      fit <- tibble::rownames_to_column(fit, "method")
      for (j in nrow(fit)) {
        scaled <- SDT2diffusion(parameters = unlist(fit[j, c("d.W", "d.S", "s", "sv.L", "v.L")]))
        fit[c("v.W","v.S","sv.T")] <- as.list(scaled[c("v.W","v.S","sv.T")])
      }
      sub_fits[[i]] <- fit
  }

  sub_fits

}

fits <- lapply(fits, bind_rows) %>%
  bind_rows() %>%
  group_by(subject) %>%
  filter(value == min(value)) %>%
  ungroup()
```

